Modelling potential environmental impacts of science activity in Antarctica

### 2024-12-01 paper/paper.qmd
latest changes to paper

---
title: Modelling potential environmental impacts of science activity in Antarctica
author: 
  - name: David O’Sullivan
    email: david.osullivan@auckland.ac.nz
    affiliations:
      - name: University of Auckland
        url: https://profiles.auckland.ac.nz/david-osullivan
      - name: Geospatial Stuff 
        url: https://dosull.github.io
  - name: Fraser J. Morgan
    affiliations: 
      - name: Manaaki Whenua Landcare Research 
        url: https://landcareresearch.co.nz
bibliography: antarctica.bib
csl: the-canadian-geographer.csl
format: 
  html:
    toc: true
    number-sections: true
    number-depth: 3
    self-contained: true
  pdf:
    documentclass: article
    classoption: [11pt]
    toc: false
    number-sections: true
    number-depth: 3
execute:
  echo: false
  warning: false
  results: false
  message: false
---

### Abstract {.unnumbered}
We use GPS data collected on a science expedition in Antarctica to estimate hiking functions for the speed at which humans traverse terrain differentiated by slope and by ground cover (moraines and rock). We use the estimated hiking functions to build weighted directed graphs as a representation of specific environments in Antarctica. From these we estimate using a variety of graph metrics—particularly betweennness centrality—the relative potential for human environmental impacts arising from scientific activities in those environments. We also suggest a simple approach to planning science expeditions that might allow for reduced impacts in these environments.

## Introduction
Overview of Antarctic science: when and where, its intensity etc. Background on international treaties, etc.

@@

## Our approach and related work
We chose to explore the question of where human impacts are likely to be strongest using an approach closely related to work on patterns of human movement in archaeology [@Verhagen2019] where likely and potential movement paths of humans across landscapes have been used to infer the settlement structure and human geography of large-scale landscapes. 
Closely related work in biology investigates the structure and geography of animal transportation networks [@Perna2014]. 
Both approaches rely on the idea that humans or animals move around in an environment in time or energy efficient ways. 
These approaches rely on hiking functions that relate speed of movement across a terrain to its slope.

@@

### Hiking functions
@Prisner2023 provide an overview of a variety of functions that have been used to model how hiking times and speeds vary with terrain slope. 
They consider longstanding rules of thumb [@Naismith1892], and later modifications thereof [@Langmuir1984], along with more recent such guidance from the Swiss and German Alpine Clubs [@Winkler2010;@Deutscher2011]. 
These functions estimate the time taken to travel 1km, referred to as _pace_, based on slope expressed as _rise over run_, that is change in elevation divided by horizontal distance. 
They are all piecewise functions with sharp changes in estimated pace at specific slopes.

Alongside these hiking pace functions @Prisner2023 also present hiking speed functions (generally referred to as simply _hiking functions_) from Tobler [-@Tobler1993 generally considered the first hiking function] and more recent, related but more firmly empirically grounded alternatives offered by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019. Another hiking function not discussed by @Prisner2023 is presented by @Rees2004. 
These hiking functions are all continuous in the slope of the terrain so that $v=f(\theta)$, where $v$ is the speed, and $\theta$ is the slope. 
They can all be parameterised to control the maximum speed attainable, the slope at which maximum speed is attained (expected to be a shallow downhill slope), and the rate at which speed falls off with increasing slope.

@@

| Description | Equation   | Examples
|:------------|------------|:--------
|Exponential  | $ae^{b\left|\theta-c\right|}$ | @Tobler1993, @Marquez-Perez2017
| Gaussian    | $ae^{-b(\theta-c)^2}$ | @Irmischer2018, @Campbell2019
| Lorentz     | $\frac{a}{\left[b+d(\theta-c)^2\right]}$ | @Campbell2019
| Quadratic   | $a+b\theta+c\theta^2$ | @Rees2004

: Functional forms of hiking functions {#tbl-hiking-function-equations .striped tbl-colwidths="[20,20,80]"}

```{r}
#| label: fig-hiking-functions
#| fig-cap: Example hiking functions.

library(ggplot2)
library(cols4all)
library(tidyr)
library(dplyr)

theme_set(theme_minimal())

tobler <- function(theta, a = 6, b = 3.5, c = -0.05) {
  a * exp(-b * abs(theta - c))
}
gaussian <- function(theta, a = 5.5, b = 0.35, c = -0.04) {
  a * exp(-(theta - c) ^ 2 / b ^ 2)
}

lorentz <- function(theta, a = 5.25, b = 1, c = -0.06, d = 10) {
  a / (b + d * (theta - c) ^ 2)
}

rees <- function(theta, a = 5.25, b = -1.5, c = -16) {
  5 + b * theta + c * theta ^ 2
}

results <- tibble(Slope = -50:50/100,
                  Exponential = tobler(Slope), 
                  Gaussian = gaussian(Slope),
                  Lorentz = lorentz(Slope),
                  Rees = rees(Slope)) |>
  pivot_longer(cols = -1) |>
  rename(Function = name, `Speed, km/h` = value)

ggplot(results) +
  geom_line(aes(x = Slope, y = `Speed, km/h`, 
            colour = Function, group = Function), linewidth = 1) +
  scale_colour_discrete_c4a_cat("brewer.set2") +
  scale_x_continuous(limits = c(-0.5, 0.5), expand = c(0, 0)) +
  theme(panel.border = element_rect(linewidth = 0.5, fill = NA))
```

The parameterisation of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the Exponential hiking function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
These parameter values for the exponential functional form give _Tobler's hiking function_ but, as has been noted elsewhere, [see @Herzog2010;@Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds presented by Imhof [-@Imhof1950, pp. 217-220].
Nevertheless these parameter values are widely applied in the literature.

All these hiking function forms are somewhat _ad hoc_. 
They exhibit desirable, expected properties: a peak speed at a slope near zero, which we expect to be slightly negative (i.e., downhill), and continuously falling speeds as the slope deviates from the slope of peak speed.
However, there is no theoretical basis for the specific functional forms listed in @tbl-hiking-function-equations.
More principled approaches might be developed based on the literature on the physiology of human movement, see e.g. **CITATIONS NEEDED HERE**.
In general, approaches based on minimising energy use yield similar results to empirical speed-slope functions, although it is worth noting that they more reliably generate zig-zag or 'switchback' movement behaviour on steep slopes [@Llobera2007].
However, these approaches are hard to implement, and empirically-derived and locally-specific hiking functions as reported by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019 are the approach we have taken in the present work, based on available data and the goals of our study where the relative cost of different potential routes in the landscape is more important than exact prediction of routes.
In practice, almost any function with the peaked form of those shown in @fig-hiking-functions is suited to our requirements.

It is commonplace in many applications to also incorporate a penalty on movement contingent on landcover, especially for off-track or off-road movement.
For example the speed attainable off-track in forested terrain might be only half that attainable in grasslands.
Unsurprisingly, there are no widely agreed landcover penalties, but see for example, those compiled by @Herzog2020.
The hiking function reported by @Wood2023 includes the local gradient of the terrain (i.e., the maximum slope at each location, not the slope in the direction of movement) as a covariate in the estimated function. 
It is possible that this kind of approach treating landcover as a categorical covariate in the estimation of more complex hiking functions might be applied in future work. 
In our application because there are only two kinds of navigable landcover&mdash;moraine (or gravel) and rock&mdash;we chose instead to estimate two hiking functions, one for each landcover and estimate movement costs conditional on the landcover present at each location.
This has the additional advantage of allowing for distinctly different effects of slope on speed.
Details of this approach are reported and discussed in @sec-hiking-function-estimation.

### Representing the landscape
Terrain is asymmetric with respect to movement, therefore: landscapes as graphs

Edge costs in graphs modelled based on slope and a hiking function

Movement cost for graph edges derived from hiking functions

Graph vertices augmented with landcover information (in this context limited to surface geology)

@@

[FIGURE showing location information etc., based on these data]

### GPS data from an expedition

Devices used, and associated protocols for scientists while on site.

[FIGURE showing GPS traces in context]

@@

[FIGURE showing filtering steps and impact on slope vs speed scatterplot/boxplots]

## Methods and results 
### Hiking functions {#sec-hiking-function-estimation}
Fit three alternative functional forms: exponential Tobler (1993, pp 1–4, based on data from Imhof 1950), Gaussian Irmischer and Clarke (2018), and Student’s t distribution, the last of these as a heavier-tailed alternative to Gaussian. Other approaches rely on rule-of-thumb hiking time calculations such as...

@@

[FIGURE showing hiking function fits about here]

@@

[FIGURE comparing moraine and rock hiking functions about here]

### Landscapes as graphs
Assignment of nodes as a lattice of equally spaced locations. Choice of rectangular/square grid vs. hexagonal. Square grids are easily generated and also offer some computational advantages if, for example, the grid is aligned with an underlying digital elevation model (DEM). However, square grids present the problem of choosing between two different potential local neighbourhoods, based on orthogonal adjacency only, or on orthogonal and diagonal adjacency (respectively referred to as rook and queen’s case adjacency, or as von Neumann and Moore neighbourhoods, in the spatial analysis and cellular automaton literatures). Excluding diagonal adjacency yields implausible diamond shaped isochrones on level terrain. Including diagonal adjacency can lead to seemingly contradictory results when alternate corners of a square are at different heights.

Given these challenges we adopt a hexagonally arranged triangular lattice. This avoids any difficulties with varying distances (i.e. network edge lengths) between nodes. Note that this is necessarily an approximation, and on extensive flat terrains will yield unrealistic hexagonal isochrones. It also presents some computational disadvantages in assigning elevations to graph nodes where interpolation of heights is required due to mismatches between hexagon centres and cells in grid-based DEMs.

[FIGURE showing square and hexagonal lattices and issues]

Alternative approaches might estimate movement based on generating multiple random approximately evenly spaced nodes across a terrain and combining the resulting estimated travel times [see @Etherington2012], or deploying methods for estimating geodesics on complex triangular meshes [@Martinez2005] but both approaches are more computationally demanding than the current application justifies. In any case, the latter method does not allow for asymmetry in traversal times as arising due to slopes.

We developed R code [@RCoreTeam2024] to build graphs (i.e. networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions. Graphs are stored as igraph package [@csardi2006igraph;@Csardi2024] graph objects for further analysis.

@@

`igraph` implements betweenness centrality measures [@Freeman1978;@Brandes2001], which we used to identify vertices and/or edges most likely to be traversed by minimum travel time routes across the terrain.

@@

### Impact minimizing networks
Tentative proposal for impact minimizing networks based on minimum spanning trees, but noting the issue with respect to directed graphs when these would more correctly be arborescences [@Korte2018a].

## Discussion
Blah

## Conclusions
Blah

>>>>

---
title: Modelling potential environmental impacts of science activity in Antarctica
author: 
  - name: David O’Sullivan
    email: david.osullivan@auckland.ac.nz
    affiliations:
      - name: University of Auckland
        url: https://profiles.auckland.ac.nz/david-osullivan
      - name: Geospatial Stuff 
        url: https://dosull.github.io
  - name: Fraser J. Morgan
    affiliations: 
      - name: Manaaki Whenua Landcare Research 
        url: https://landcareresearch.co.nz
bibliography: antarctica.bib
csl: the-canadian-geographer.csl
lightbox: true
format: 
  html:
    toc: true
    number-sections: true
    number-depth: 3
    self-contained: true
  pdf:
    documentclass: article
    classoption: [11pt]
    toc: false
    number-sections: true
    number-depth: 3
    linestretch: 1.15
execute:
  echo: false
  warning: false
  results: false
  message: false
---
```{css, echo = FALSE}
p {color: black;}
figcaption {color: black;}
```

```{r}
library(ggplot2)
library(ggpubr)
library(grid)
library(cols4all)
library(tidyr)
library(dplyr)
library(igraph)
library(sf)
```

### Abstract {.unnumbered}

We use GPS data collected on a science expedition in Antarctica to estimate hiking functions for the speed at which humans traverse terrain differentiated by slope and by ground cover (moraines and rock). 
We use the estimated hiking functions to build weighted directed graphs as a representation of specific environments in Antarctica. 
From these we estimate using a variety of graph metrics—particularly betweennness centrality—the relative potential for human environmental impacts arising from scientific activities in those environments. 
We also suggest a simple approach to planning science expeditions that might allow for reduced impacts in these environments.

## Introduction

Overview of Antarctic science: when and where, its intensity etc. Background on international treaties, etc.

@@

## Our approach and related work

We chose to explore the question of where human impacts are likely to be strongest using an approach closely related to work on patterns of human movement in archaeology [@Verhagen2019] where likely and potential movement paths of humans across landscapes have been used to infer the settlement structure and human geography of large-scale landscapes. 
Closely related work in biology investigates the structure and geography of animal transportation networks [@Perna2014]. 
Both approaches rely on the idea that humans or animals move around in an environment in time or energy efficient ways. 
These approaches rely on hiking functions that relate speed of movement across a terrain to its slope.

@@

### Hiking functions

@Prisner2023 provide an overview of a variety of functions that have been used to model how hiking times and speeds vary with terrain slope. 
They consider longstanding rules of thumb [@Naismith1892], and later modifications thereof [@Langmuir1984], along with more recent such guidance from the Swiss and German Alpine Clubs [@Winkler2010; @Deutscher2011]. 
These functions estimate the time taken to travel 1km, referred to as *pace*, based on slope expressed as *rise over run*, that is change in elevation divided by horizontal distance. 
They are all piecewise functions with sharp changes in estimated pace at specific slopes.

Alongside these hiking pace functions @Prisner2023 also present hiking speed functions (generally referred to as simply *hiking functions*) from Tobler [-@Tobler1993 generally considered the first hiking function] and more recent, related but more firmly empirically grounded alternatives offered by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019. 
Another hiking function not discussed by @Prisner2023 is presented by @Rees2004. 
These hiking functions are all continuous in the slope of the terrain so that $v=f(\theta)$, where $v$ is the speed, and $\theta$ is the slope. 
They can all be parameterised to control the maximum speed attainable, the slope at which maximum speed is attained (expected to be a shallow downhill slope), and the rate at which speed falls off with increasing slope.

@@

| Description | Equation                                 | Examples                        |
|:--------------------------|-------------------------|:------------------|
| Exponential | $ae^{b\left|\theta-c\right|}$            | @Tobler1993, @Marquez-Perez2017 |
| Gaussian    | $ae^{-b(\theta-c)^2}$                    | @Irmischer2018, @Campbell2019   |
| Lorentz     | $\frac{a}{\left[b+d(\theta-c)^2\right]}$ | @Campbell2019                   |
| Quadratic   | $a+b\theta+c\theta^2$                    | @Rees2004                       |

: Functional forms of hiking functions {#tbl-hiking-function-equations .striped tbl-colwidths="\[20,20,80\]"}

```{r}
#| label: fig-hiking-functions
#| fig-cap: "Example hiking functional forms: Exponential [@Tobler1993], Gaussian [@Irmischer2018], Lorentz [@Campbell2019], and Quadratic [@Rees2004]."

theme_set(theme_minimal())
theme_update(panel.border = element_rect(linewidth = 0.5, fill = NA))

tobler <- function(theta, a = 6, b = 3.5, c = -0.05) {
  a * exp(-b * abs(theta - c))
}
gaussian <- function(theta, a = 5.5, b = 0.25, c = -0.04) {
  a * exp(-(theta - c) ^ 2 / 2 / b ^ 2)
}
lorentz <- function(theta, a = 4.75, b = 0.9, c = -0.06, d = b / pi) {
  a / (b * (1 + ((theta - c) / d) ^ 2))
}
rees <- function(theta, a = 5.25, b = -1.5, c = -16.5) {
  5 + b * theta + c * theta ^ 2
}

results <- tibble(Slope = -50:50/100,
                  Exponential = tobler(Slope), 
                  Gaussian = gaussian(Slope),
                  Lorentz = lorentz(Slope),
                  Quadratic = rees(Slope)) |>
  pivot_longer(cols = -1) |>
  rename(`Functional form` = name, `Speed, km/h` = value)

ggplot(results) +
  geom_line(aes(x = Slope, y = `Speed, km/h`, linetype = `Functional form`,
            colour = `Functional form`, group = `Functional form`), linewidth = 1) +
  scale_colour_discrete_c4a_cat("brewer.set1") +
  scale_linetype_manual(values = c("solid", "longdash", "dotdash", "dotted")) + 
  scale_x_continuous(limits = c(-0.5, 0.5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 6.25), expand = c(0, 0)) +
  geom_vline(aes(xintercept = -0.05), lwd = 0.35, linetype = "dotted") +
  annotate("label", x = -0.055, y = 0.5, label = "Peak speed slope (exponential)", 
           angle = 90, hjust = 0, vjust = 0.6, label.size = 0)
```

The parameterisations of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the exponential function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993. 
These parameters give *Tobler's hiking function* but, as has been noted elsewhere, [see @Herzog2010; @Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds presented by Imhof [-@Imhof1950, pp. 217-220]. 
Nevertheless these parameter values are widely applied in the literature.

All these hiking functional forms are somewhat *ad hoc*. 
They exhibit desirable, expected properties: a peak speed at a slope near zero, which we expect to be slightly negative (i.e., downhill), and continuously falling speeds as the slope deviates from the slope of peak speed. 
However, there is no theoretical basis for the specific functional forms listed in @tbl-hiking-function-equations. 
More principled approaches might be developed based on the literature on the physiology of human movement, see e.g. **CITATIONS NEEDED HERE**. 
In general, approaches based on minimising energy use yield similar results to empirical speed-slope functions, although it is worth noting that they more reliably generate zig-zag or 'switchback' movement behaviour on steep slopes [@Llobera2007]. 
However, these are hard to implement, and we have adopted empirically-derived and locally-specific hiking functions following @Marquez-Perez2017, @Irmischer2018, and @Campbell2019. 
This choice is based on available data and the goals of our study, where the relative cost of different potential routes in the landscape is more important than exact prediction of routes. 
In practice, almost any function with the peaked form of those shown in @fig-hiking-functions is suited to our requirements.

It is commonplace in many applications to also incorporate a penalty on movement contingent on landcover, especially for off-track or off-road movement. 
For example the speed attainable off-track in forested terrain might be only half that attainable in grasslands. 
Unsurprisingly, there are no widely agreed landcover penalties, but see for example, those compiled by @Herzog2020. 
A hiking function derived by @Wood2023 includes the local gradient of the terrain (i.e., the maximum slope at each location, not the slope in the direction of movement) as a covariate in the estimated function. 
It is possible that this kind of approach treating landcover as a categorical covariate in the estimation of more complex hiking functions might be more widely applied in future work. 
In our application, because there are only two kinds of navigable landcover&mdash;moraine (or gravel) and rock&mdash;we chose instead to estimate two hiking functions, one for each landcover, and estimate movement costs conditional on the landcover present at each location. 
This also has the advantage of allowing for differing effects of slope on speed due to landcover, where for example, gravel might allow more rapid movement on the level surfaces, but more rapidly reduce speed on slopes. 
Details of this approach are reported and discussed in @sec-hiking-function-estimation.

### Representing the landscape
Hiking functions are usually asymmetric with respect to slope, with the highest attainable speed at a slight downhill slope. 
This asymmetry means that unless analysis is focused on assessing movement cost from a single origin or to a single destination, it is necessary to represent the landscape in a way that can accommodate asymmetry. 
We therefore represent the landscape as a directed graph. 
Graph vertices are locations with associated elevation and landcover. 
Graph edges are directed links between vertices for which a change in elevation between the start and end vertex can be calculated, and a slope derived, based on the elevation difference and horizontal distance between the vertices.

The most obvious way to derive such a graph is to assign each cell in a raster digital elevation model (DEM) to a graph vertex, so that graph vertices are arranged in a regular square grid or lattice. 
This approach has been widely adopted **CITATIONS NEEDED HERE** 
However, this still requires a decision to be made regarding the definition of graph edges, which can be thought of as _allowed moves_ between vertices. 
Three possibilities on a square lattice&mdash;Rook's, Queen's, and Knight's moves&mdash;are shown in @fig-lattice-adjacencies. Another option shown in the figure is to lay out a regular hexagonal grid of vertex locations and link each vertex with its six nearest neighbours.

```{r}
#| label: fig-lattice-adjacencies
#| fig-cap: Possible graph lattices.
#| fig-width: 7
#| fig-height: 7

theme_set(theme_void()) 
theme_update(plot.margin = unit(c(1, 1, 2, 1), "mm"),
             plot.title = element_text(hjust = 0.5, vjust = 2))

get_graph_adjacencies <- function(G) {
  edges <- ends(G, E(G))
  list(i = edges[, 1], j = edges[, 2])
}

make_linestring <- function(x1, y1, x2, y2) {
  st_linestring(matrix(c(x1, y1, x2, y2), ncol = 2, byrow = TRUE))
}

get_graph_as_line_layer <- function(G, crs) {
  ij <- get_graph_adjacencies(G)
  i <- ij$i
  j <- ij$j
  edges_as_sf <- mapply(make_linestring, 
                        V(G)$x[i], V(G)$y[i], V(G)$x[j], V(G)$y[j],
                        SIMPLIFY = FALSE) |>
    st_sfc() |>
    st_as_sf(crs = crs) |>
    rename(geometry = x)
  edges_as_sf                      
}

get_lattice <- function(dist, pts) {
  pts |>
    st_buffer(dist) |>
    st_contains(pts, sparse = FALSE) |>
    as.matrix() |>
    graph_from_adjacency_matrix(mode = "undirected", diag = FALSE) |>
    set_vertex_attr("x", value = pts$x) |>
    set_vertex_attr("y", value = pts$y)
}

get_graph_plot <- function(G, main, crs = 2193) {
  return(ggplot() +
    geom_sf(data = G |> 
              get_graph_as_line_layer(crs = 2193)) +
    geom_sf(data = G |> 
              vertex_attr() |> as.data.frame() |> st_as_sf(coords = 1:2, crs = crs), 
            size = 8, fill = "white", colour = "black", shape = 21, lwd = 1) +
    coord_sf(xlim = c(-270, 270) + 1e6, 
             ylim = c(-270, 270) + 1e6, expand = FALSE) +
    ggtitle(main))
}

grid_pts <- expand_grid(x = -5:5 * 100 + 1e6, y = -5:5 * 100 + 1e6) |>
  as.data.frame() |>
  st_as_sf(coords = c("x", "y"), crs = 2193, remove = FALSE)

hex_grid_pts <- grid_pts |>
  st_bbox() |> 
  st_make_grid(square = FALSE, what = "centers", cellsize = 107.457) |>
  st_as_sf() |>
  rename(geometry = x) |>
  mutate(geometry = geometry + c(-9, -11)) |>
  st_set_crs(2193)

xy <- hex_grid_pts |> st_coordinates()
hex_grid_pts <- hex_grid_pts |>
  mutate(x = xy[, 1], y = xy[, 2])

graphs <- mapply(
  get_lattice,
  list(101, 150, 225, 110),
  list(grid_pts, grid_pts, grid_pts, hex_grid_pts))
names(graphs) <- list("Rook's move", "Queen's move", "Knight's move", "Hex")

ggplots <- list()
for (n in names(graphs)) {
  ggplots[[n]] <- get_graph_plot(G = graphs[[n]], main = n)
}
ggarrange(plotlist = ggplots, ncol = 2, nrow = 2)
```

The advantage of a square lattice is the ease of deriving vertex elevations assuming a DEM at the desired lattice spacing is available. 
A hexagonal lattice on the other hand will usually require that elevation values be interpolated from a DEM to the lattice locations.

Both square and hexagonal lattices can lead to geometric artifacts when the allowed moves are used to determine contours of equal travel time or *isochrones* from an origin point. 
On level terrain the Rook's move produces diamond-shaped isochrones, while the Queen's move produces octagons. 
In general isochrones will be polygons with as many sides as the number of neighbours of each vertex. 
While the 16-gons of the Knight's move lattice are likely to be close enough to circular for many purposes, the resulting graph is much larger than the alternatives. @Etherington2012 suggests that by combining results from randomly generated planar graphs such geometric artifacts can be removed, but this substantially increases the computational requirements. Another computationally intensive approach might draw on methods for
 estimating geodesics on complex triangular meshes [@Martinez2005].
Because the geometric effects of the graph structure are masked when we introduce varying movement speeds due to the slope of each edge, we do not consider these more complex approaches necessary in our application. 
Nevertheless, more circular base isochrone shapes are to be preferred, all other things equal.

A further complication when a lattice includes edges of varying length as in the Queen's and Knight's move lattices, is bias in the estimation of slopes leading to lower estimated movement costs for longer edges. 
This is because longer edges (e.g., the knight's moves) may 'jump' across intervening segments of varying slope, smoothing them to a single slope estimate.
Such smoothing will usually result in estimated speeds along that edge higher than those that would accumulate along the intervening sections of varying slope. 
This problem is discussed in a slightly different context by Campbell et al. [-@Campbell2019 pp. 96-98].

Any choice of graph structure is necessarily a compromise, and estimates of movement rates are always an approximation [@Goodchild2020]. 
We consider the differing edge lengths introduced by all the square lattices other than the Rook's case to be problematic, and favour the hexagonal lattice structure over the simple square lattice because a base hexagonal isochrone is preferable to a diamond shape.

Having settled on a hexagonal lattice for the graph structure, at a chosen resolution we set out a regular hexagonal grid of locations across the study area, and assign to each point an elevation by interpolation from a DEM. 
Because the DEM is at finer resolution than the hexagonal lattice, the choice of interpolation method is not a major concern.
Our approach applies a bilinear interpolation method based on elevation values in the DEM cell the graph vertex falls in and its four orthogonal neighbours.
Based on the difference in elevation of the vertices at each end of each edge we estimate a slope, and also a traversal times using our estimated hiking function. 
The important point here is that our graph is _directed_ so that different traversal times $t_{ij}$ and $t_{ji}$ are estimated for moving between locations $i$ to location $j$ depending on the direction of movement.

As noted in the previous section, we estimate two hiking functions, one for moraine and one for rock landcover. 
When an edge connects locations of two different landcover types the estimated traversal time is the mean of the traversal times for each landcover.

@@

\[FIGURE showing location information etc., based on these data\]

### GPS data from an expedition
Devices used, and associated protocols for scientists while on site.

\[FIGURE showing GPS traces in context\]

@@

\[FIGURE showing filtering steps and impact on slope vs speed scatterplot/boxplots\]

## Methods and results

### Hiking functions {#sec-hiking-function-estimation}
Fit three alternative functional forms: exponential Tobler (1993, pp 1–4, based on data from Imhof 1950), Gaussian Irmischer and Clarke (2018), and Student’s t distribution, the last of these as a heavier-tailed alternative to Gaussian.
Other approaches rely on rule-of-thumb hiking time calculations such as...

@@

\[FIGURE showing hiking function fits about here\]

@@

\[FIGURE comparing moraine and rock hiking functions about here\]

### Landscapes as graphs
We developed R code [@RCoreTeam2024] to build graphs (i.e. networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions. Graphs are stored as igraph package [@csardi2006igraph; @Csardi2024] graph objects for further analysis.

@@

`igraph` implements betweenness centrality measures [@Freeman1978; @Brandes2001], which we used to identify vertices and/or edges most likely to be traversed by minimum travel time routes across the terrain.

@@

### Impact minimizing networks

Tentative proposal for impact minimizing networks based on minimum spanning trees, but noting the issue with respect to directed graphs when these would more correctly be arborescences [@Korte2018a].

## Discussion

Blah

## Conclusions

Blah
